---
title: "ANA 515 Assignment 4"
author: Alexander Tyan, Jie Hui Ho
date: March 12, 2023
output: pdf_document
---

## 1. Discuss the business problem/goal

The goals of this project is to create a recommendation engine that recommends 
films that appeal to users based on their preferences and browsing history. The
 recommendation are powered by a Machine Learning algorithm.

## 2. identify where the dataset was retrieved from

The MovieLens dataset was retrieved from  
`https://drive.google.com/file/d/1Dn1BZD3YxgBQJSIjbfNnmCFlDW2jdQGD/view`  
which has two csv files – `movies.csv` and `ratings.csv`.

## 3. Identify the code that imported and saved your dataset in R

```{r, echo = FALSE, results = "hide", include = FALSE}
library(tidyverse)
library(recommenderlab)
library(data.table)
library(ggplot2)
library(reshape2)
```

``` {r}
# Reading in the dataset, as downloaded from the links provided at
# https://data-flair.training/blogs/data-science-r-movie-recommendation/
# and saved into personal GitHub space:
movie_data <- read_csv(
    "https://raw.githubusercontent.com/dr3am05/Source/main/movies.csv"
)
rating_data <- read_csv(
    "https://raw.githubusercontent.com/dr3am05/Source/main/ratings.csv"
)
```

## 4. Describe your data set 



This dataset consists of `r nrow(rating_data)` ratings applied over 
`r nrow(movie_data)` movies. Movie data has `r nrow(movie_data)` rows with 
`r ncol(movie_data)` columns, named `movieId`, `title` and `genres`. `movieId` is 
an integer type while `title` and `genres` are character type.
Rating data has `r nrow(rating_data)` rows with `r ncol(rating_data)` 
columns, named `userId`, `movieId`, `rating` and `timestamp`.

Below is a summary of our movie data and ratings data by each variable. 

``` {r, include = TRUE, echo=TRUE}
summary(movie_data)
summary(rating_data)
```

The movie data has `r sum(is.na(movie_data))` missing values; 
and the ratings data has `r sum(is.na(rating_data))` missing values.

## 5. Discuss any data preparation, missing values and errors

As discussed in 4, there are no missing values in both datasets

Because Machine Learning needs data represented in matrix forms, we need to
prepare our datasets into that format.

First, to train a model, we convert the movie data into a sparse matrix of 
`0`'s and `1`'s, where each column is a movie genre, and each row is a movie
title. This matrix is `genre_mat2` and we can see its truncated version in the
output below (though `str()` transposes the matrix visually, flipping
movie titles/genres):
```{r}
movie_genre <- as.data.frame(movie_data$genres, stringsAsFactors = FALSE)
movie_genre2 <- as.data.frame(
    tstrsplit(movie_genre[, 1], "[|]",
        type.convert = TRUE
    ),
    stringsAsFactors = FALSE
)
colnames(movie_genre2) <- c(1:10)

list_genre <- c(
    "Action", "Adventure", "Animation", "Children",
    "Comedy", "Crime", "Documentary", "Drama", "Fantasy",
    "Film-Noir", "Horror", "Musical", "Mystery", "Romance",
    "Sci-Fi", "Thriller", "War", "Western"
)
genre_mat1 <- matrix(0, 10330, 18)
genre_mat1[1, ] <- list_genre
colnames(genre_mat1) <- list_genre

for (index in 1:nrow(movie_genre2)) {
    for (col in 1:ncol(movie_genre2)) {
        gen_col <- which(genre_mat1[1, ] == movie_genre2[index, col])
        genre_mat1[index + 1, gen_col] <- 1
    }
}
# remove first row, which was the genre list
genre_mat2 <- as.data.frame(genre_mat1[-1, ], stringsAsFactors = FALSE)
for (col in 1:ncol(genre_mat2)) {
    genre_mat2[, col] <- as.integer(genre_mat2[, col]) # convert from characters to integers
}
str(genre_mat2)
```

Next, we create a search matrix that merge `movie_data` and `genre_mat2`. 
This allows user to easily perform a search of the movie titles based on 
the genres listed in our list. This merged matrix is `SearchMatrix` below.

```{r}
# Using column bind function to merge movie_data and genre_mat2 dataframes
SearchMatrix <- cbind(movie_data[, 1:2], genre_mat2[])
head(SearchMatrix)
```

Then, we create a sparse matrix of ratings. Each row represents a user and each
column is a movie. That makes each entry in the matrix (`ratingMatrix`) a
rating given by a particular user for a particular movie. The matrix is sparse 
because most movies are not rated by most users. The output below shows the
matrix dimensions. We force that object to be of `realRatingMatrix` type, for
use by our `recommenderlab` Machine Learning package later.

```{r}
ratingMatrix <- dcast(
    rating_data, userId ~ movieId,
    value.var = "rating", na.rm = FALSE
)
ratingMatrix <- as.matrix(ratingMatrix[, -1]) # remove userIds
# Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix
```

We now filter out data to only keep what is useful. Let us only keep 
that have "enough" data. "Enough" in this case is films rated by more than 50
users and users who have rated more than 50 films. 
```{r}
movie_ratings <- ratingMatrix[
    rowCounts(ratingMatrix) > 50,
    colCounts(ratingMatrix) > 50
]
movie_ratings
```

We can visualize top users/top movies with this heatmap; the darker the square,
the higher the rating given by that particular user (indicated by a row) given
to that particular movie (given by a column) TODO: (might move to 8):
```{r}
minimum_movies <- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(
    movie_ratings[
        rowCounts(movie_ratings) > minimum_movies,
        colCounts(movie_ratings) > minimum_users
    ],
    main = "Heatmap of the top users and movies"
)
```
Let us visualize distribution of the average ratings per user:
TODO: (might move to 8)
```{r}
average_ratings <- rowMeans(movie_ratings)
qplot(
    average_ratings,
    fill = I("steelblue"),
    col = I("red")
) +
    ggtitle("Distribution of the average rating per user")
```

The next step in data preparation is to normalize user ratings to have mean of 
0 to limit bias of extreme ratings on the model training later:
```{r}

normalized_ratings <- normalize(movie_ratings)
sum(rowMeans(normalized_ratings) > 0.00001)
image(
    normalized_ratings[
        rowCounts(normalized_ratings) > minimum_movies,
        colCounts(normalized_ratings) > minimum_users
    ],
    main = "Normalized Ratings of the Top Users"
)
```

As our final step in data preparation, we will binarize this data to have 
discrete values of 0 and 1. This allows us to create a matrix that will
consist of 1 if the movie rating is above 3 and 0 otherwise. This will allow
for a more efficient recommendation system later; this is also visualized:
```{r}
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
# movies_watched <- binarize(movie_ratings, minRating = 1)

good_rated_films <- binarize(movie_ratings, minRating = 3)
image(
    good_rated_films[
        rowCounts(movie_ratings) > binary_minimum_movies,
        colCounts(movie_ratings) > binary_minimum_users
    ],
    main = "Heatmap of the top users and movies"
)
```

## 6. Discuss the modeling. What modeling was used?

We will first split our dataset into 80% training set `training_data`
and 20% test set `testing_data`

```{r}
# splitting the dataset to 80% training set and 20% test set
sampled_data <- sample(
    x = c(TRUE, FALSE),
    size = nrow(movie_ratings),
    replace = TRUE,
    prob = c(0.8, 0.2)
)
training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
```

For the modeling part of this project, we will be utilizing Item Based 
Collaborative Filtering System. This type of collaborative filtering looks for
similarities between items based on the movie ratings.
Since the parameters are default by nature, we will set `k` parameter to 30. 
`k` represents the number of items for computing similarities, the algorithm 
will identify `k` most similar items and store their respective number.

```{r}
recommen_model <- Recommender(
    data = training_data,
    method = "IBCF",
    parameter = list(k = 30)
)
recommen_model
class(recommen_model)
```

After we retrieve the `recommen_model`, we then generate a heatmap that contains top 20

```{r}
model_info <- getModel(recommen_model)
class(model_info$sim)
dim(model_info$sim)
top_items <- 20
image(model_info$sim[1:top_items, 1:top_items],
   main = "Heatmap of the first rows and columns")
```

```{r}
sum_rows <- rowSums(model_info$sim > 0)
table(sum_rows)

sum_cols <- colSums(model_info$sim > 0)
qplot(sum_cols, fill=I("steelblue"), col=I("red"))+ ggtitle("Distribution of the column count")
```

```{r}
top_recommendations <- 10 # the number of items to recommend to each user
predicted_recommendations <- predict(object = recommen_model,
                                     newdata = testing_data,
                                     n = top_recommendations)
predicted_recommendations
```

```{r}
user1 <- predicted_recommendations@items[[1]] # recommendation for the first user
movies_user1 <- predicted_recommendations@itemLabels[user1]
movies_user2 <- movies_user1
for (index in 1:10){
  movies_user2[index] <- as.character(subset(movie_data,
                                         movie_data$movieId == movies_user1[index])$title)
}
movies_user2
```

```{r}
recommendation_matrix <- sapply(predicted_recommendations@items,
                      function(x){ as.integer(colnames(movie_ratings)[x]) }) # matrix with the recommendations for each user
#dim(recc_matrix)
recommendation_matrix[,1:4]
```

movie_data %>% filter(movieId == 145)

## 8. Visualization