---
title: "ANA 515 Assignment 4"
author: Alexander Tyan, Jie Hui Ho
date: March 12, 2023
output: pdf_document
---

## 1. Discuss the business problem/goal

The goals of this project is to create a recommendation engine that recommends 
films that appeal to users based on their preferences and browsing history. The
 recommendation are powered by a Machine Learning algorithm.

## 2. identify where the dataset was retrieved from

The MovieLens dataset was retrieved from  
`https://drive.google.com/file/d/1Dn1BZD3YxgBQJSIjbfNnmCFlDW2jdQGD/view`  
which has two csv files – `movies.csv` and `ratings.csv`.

## 3. Identify the code that imported and saved your dataset in R

```{r, echo = FALSE, results = "hide", include = FALSE}
library(tidyverse)
library(recommenderlab)
library(data.table)
library(ggplot2)
library(reshape2)
```

``` {r}
# Reading in the dataset, as downloaded from the links provided at
# https://data-flair.training/blogs/data-science-r-movie-recommendation/
# and saved into personal GitHub space:
movie_data <- read_csv(
    'https://raw.githubusercontent.com/dr3am05/Source/main/movies.csv'
)
rating_data <- read_csv(
    'https://raw.githubusercontent.com/dr3am05/Source/main/ratings.csv'
)
```

## 4. Describe your data set 



This dataset consists of `r nrow(rating_data)` ratings applied over 
`r nrow(movie_data)` movies. Movie data has `r nrow(movie_data)` rows with 
`r ncol(movie_data)` columns, named `movieId`, `title` and `genres`. `movieId` is 
an integer type while `title` and `genres` are character type.
Rating data has `r nrow(rating_data)` rows with `r ncol(rating_data)` 
columns, named `userId`, `movieId`, `rating` and `timestamp`.

Below is a summary of our movie data and ratings data by each variable. 

``` {r, include = TRUE, echo=TRUE}
summary(movie_data)
summary(rating_data)
```

The movie data has `r sum(is.na(movie_data))` missing values; 
and the ratings data has `r sum(is.na(rating_data))` missing values.

## 5. Discuss any data preparation, missing values and errors

As discussed in 4, there are no missing values in both datasets

Because Machine Learning needs data represented in matrix forms, we need to
prepare our datasets into that format.

First, to train a model, we convert the movie data into a sparse matrix of 
`0`'s and `1`'s, where each column is a movie genre, and each row is a movie
title. This matrix is `genre_mat2` and we can see its truncated version in the
output below (though `str()` transposes the matrix visually, flipping
movie titles/genres):
```{r}
movie_genre <- as.data.frame(movie_data$genres, stringsAsFactors=FALSE)
movie_genre2 <- as.data.frame(tstrsplit(movie_genre[,1], '[|]', 
                                   type.convert=TRUE), 
                         stringsAsFactors=FALSE)
colnames(movie_genre2) <- c(1:10)

list_genre <- c("Action", "Adventure", "Animation", "Children", 
                "Comedy", "Crime", "Documentary", "Drama", "Fantasy",
                "Film-Noir", "Horror", "Musical", "Mystery","Romance",
                "Sci-Fi", "Thriller", "War", "Western")
genre_mat1 <- matrix(0,10330,18)
genre_mat1[1,] <- list_genre
colnames(genre_mat1) <- list_genre

for (index in 1:nrow(movie_genre2)) {
  for (col in 1:ncol(movie_genre2)) {
    gen_col = which(genre_mat1[1,] == movie_genre2[index,col])
    genre_mat1[index+1,gen_col] <- 1
}
}
#remove first row, which was the genre list
genre_mat2 <- as.data.frame(genre_mat1[-1,], stringsAsFactors=FALSE) 
for (col in 1:ncol(genre_mat2)) {
  genre_mat2[,col] <- as.integer(genre_mat2[,col]) #convert from characters to integers
}
str(genre_mat2)
```

Next, we create a search matrix that merge `movie_data` and `genre_mat2`. 
This allows user to easily perform a search of the movie titles based on 
the genres listed in our list. This merged matrix is `SearchMatrix` below.

```{r}
# Using column bind function to merge movie_data and genre_mat2 dataframes
SearchMatrix <- cbind(movie_data[, 1:2], genre_mat2[])
head(SearchMatrix)
```

Then, we create a sparse matrix of ratings. Each row represents a user and each
column is a movie. That makes each entry in the matrix (`ratingMatrix`) a
rating given by a particular user for a particular movie. The matrix is sparse 
because most movies are not rated by most users. The output below shows the
matrix dimensions. We force that object to be of `realRatingMatrix` type, for
use by our `recommenderlab` package.

```{r}
ratingMatrix <- dcast(
    rating_data, userId~movieId, value.var = "rating", na.rm = FALSE
)
ratingMatrix <- as.matrix(ratingMatrix[, -1]) #remove userIds
#Convert rating matrix into a recommenderlab sparse matrix
ratingMatrix <- as(ratingMatrix, "realRatingMatrix")
ratingMatrix
```

```{r}
movie_ratings <- ratingMatrix[
    rowCounts(ratingMatrix) > 50,
    colCounts(ratingMatrix) > 50
]
movie_ratings
```

```{r}
# Setting the threshold at 50 for the minimum number of users who have rated a
# film and minimum number of views per film
minimum_movies <- quantile(rowCounts(movie_ratings), 0.98)
minimum_users <- quantile(colCounts(movie_ratings), 0.98)
image(
    movie_ratings[
        rowCounts(movie_ratings) > minimum_movies,
        colCounts(movie_ratings) > minimum_users
    ],
        main = "Heatmap of the top users and movies"
)
```

```{r}
# Visualize distribution of the average ratings per user (might move to 8)
average_ratings <- rowMeans(movie_ratings)
qplot(
    average_ratings,
    fill = I("steelblue"),
    col = I("red")
    ) +
ggtitle("Distribution of the average rating per user")
```

```{r}
# Normalize user ratings to have mean of 0 to limit bias of extreme ratings
# in the model later:
normalized_ratings <- normalize(movie_ratings)
sum(rowMeans(normalized_ratings) > 0.00001)
image(normalized_ratings[rowCounts(normalized_ratings) > minimum_movies,
                          colCounts(normalized_ratings) > minimum_users],
main = "Normalized Ratings of the Top Users")
```

``` {r}
# Visualize distribution of normalized average ratings per user
average_ratings_normalized <- rowMeans(normalized_ratings)
qplot(
    average_ratings_normalized,
    fill = I("steelblue"),
    col = I("red")
    ) +
ggtitle("Distribution of the normalized average rating per user")
```

``` {r}
# Binarize data to discrete values of 0 and 1, create a matrix that will consist
# of 1 if the rating is above 3, otherwise 0
binary_minimum_movies <- quantile(rowCounts(movie_ratings), 0.95)
binary_minimum_users <- quantile(colCounts(movie_ratings), 0.95)
#movies_watched <- binarize(movie_ratings, minRating = 1)

good_rated_films <- binarize(movie_ratings, minRating = 3)
image(good_rated_films[rowCounts(movie_ratings) > binary_minimum_movies,
colCounts(movie_ratings) > binary_minimum_users],
main = "Heatmap of the top users and movies")
```

## 6. Discuss the modeling. What modeling was used?

```{r}
sampled_data <- sample(x = c(TRUE, FALSE),
                      size = nrow(movie_ratings),
                      replace = TRUE,
                      prob = c(0.8, 0.2))
training_data <- movie_ratings[sampled_data, ]
testing_data <- movie_ratings[!sampled_data, ]
```

## 8. Visualization